Annotated example of index standardization with VAST

### 2022-05-12 doc/appendix-pcod-VAST.Rmd
Update VAST/sdmTMB delta appendix

settings <- make_settings(
  n_x = 205, # number of vertices in the SPDE mesh
  Region = "User",
  purpose = "index2", # index of abundance with Gamma for positive catches
  fine_scale = TRUE, # use bilinear interpolation from the INLA 'A' matrix
  zone = 9,
  FieldConfig = FieldConfig,
  RhoConfig = RhoConfig,
  ObsModel = c(2, 0), # conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
  bias.correct = FALSE,
  use_anisotropy = FALSE,
  max_cells = Inf, # use all grid cells from the extrapolation grid
  knot_method = "grid" # or "samples"
)
```

@@

# effort is 1 when using CPUE instead of observed weight as the response:
pcod$effort <- 1
pcod <- as.data.frame(pcod) # ensure not a tibble
f <- here("doc", "appendix-VAST", "vast-cache.rds")
if (!file.exists(f)) {
  tictoc::tic()
  fit <- fit_model(
    settings = settings,
    Lat_i = pcod[, "lat"],
    Lon_i = pcod[, "lon"],
    t_i = pcod[, "year"],
    b_i = pcod[, "density"],
    a_i = pcod[, "effort"],
    input_grid = input_grid,
    working_dir = paste0(here("doc", "appendix-VAST"), "/")
  )
  tictoc::toc()
  saveRDS(fit, file = f)
} else {
  fit <- readRDS(f)
}
```

@@

tictoc::tic()
fit_sdmTMB <- sdmTMB(
  density ~ 0 + as.factor(year), 
  data = pcod, mesh = mesh, 
  family = delta_gamma(),
  time = "year", silent = FALSE,
  control = sdmTMBcontrol(newton_loops = 1)
)
tictoc::toc()
fit_sdmTMB
```

```{r compare-vast-sdmTMB1, echo = FALSE}
plot_betas <- function(vast_model, sdmTMB_model, vast_par = "beta1_ft", sdmTMB_pars = 1) {
  s <- vast_model$parameter_estimates$SD
  vast_est1 <- as.list(s, "Estimate", report = FALSE)
  vast_est2 <- as.list(s, "Estimate", report = TRUE)
  vast_sd1 <- as.list(s, "Std. Error", report = FALSE)
  vast_sd2 <- as.list(s, "Std. Error", report = TRUE)
  sdmTMB_est <- as.list(sdmTMB_model$sd_report, "Estimate", report = FALSE)
  sdmTMB_sd <- as.list(sdmTMB_model$sd_report, "Std. Error", report = FALSE)
  b_year_vast <- vast_est1[[vast_par]][!is.na(vast_sd1[[vast_par]])]
  b_year_vast_se <- vast_sd1[[vast_par]][!is.na(vast_sd1[[vast_par]])]
  years <- sort(unique(pcod$year))
  lwr_vast <- b_year_vast - 2 * b_year_vast_se
  upr_vast <- b_year_vast + 2 * b_year_vast_se
  plot(years, b_year_vast, ylim = range(c(lwr_vast, upr_vast)))
  segments(years, lwr_vast, years, upr_vast)
  years <- years + 0.05
  points(years, sdmTMB_est$b_j[,sdmTMB_pars])
  segments(years, sdmTMB_est$b_j[,sdmTMB_pars] - 2 * sdmTMB_sd$b_j[,sdmTMB_pars], 
    years, sdmTMB_est$b_j[,sdmTMB_pars] + 2 * sdmTMB_sd$b_j[,sdmTMB_pars],
    col = "red")
  legend("topright", legend = c("VAST", "sdmTMB"), 
    col = c("black", "red"), bty = "n", lty = c(1, 1))
}
```

>>>>

settings <- make_settings(
  n_x = 185, # number of vertices in the SPDE mesh
  Region = "User",
  purpose = "index2", # index of abundance with Gamma for positive catches
  fine_scale = TRUE, # use bilinear interpolation from the INLA 'A' matrix
  zone = 9,
  FieldConfig = FieldConfig,
  RhoConfig = RhoConfig,
  ObsModel = c(2, 0), # conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
  bias.correct = FALSE,
  use_anisotropy = FALSE,
  max_cells = Inf, # use all grid cells from the extrapolation grid
  knot_method = "grid" # or "samples"
)
```

@@

# effort is 1 when using CPUE instead of observed weight as the response:
pcod$effort <- 1
if (FALSE) { # to check offset/effort
  set.seed(1)
  pcod$effort <- exp(rnorm(nrow(pcod), mean = 0, sd = 0.2))
}
pcod <- as.data.frame(pcod) # ensure not a tibble
f <- here("doc", "appendix-VAST", "vast-cache.rds")
if (!file.exists(f)) {
  tictoc::tic()
  fit <- fit_model(
    settings = settings,
    Lat_i = pcod[, "lat"],
    Lon_i = pcod[, "lon"],
    t_i = pcod[, "year"],
    b_i = pcod[, "density"],
    a_i = pcod[, "effort"],
    input_grid = input_grid,
    working_dir = paste0(here("doc", "appendix-VAST"), "/")
  )
  tictoc::toc()
  saveRDS(fit, file = f)
} else {
  fit <- readRDS(f)
}
```

@@

tictoc::tic()
fit_sdmTMB <- sdmTMB(
  density ~ 0 + as.factor(year), 
  data = pcod, mesh = mesh,
  offset = log(pcod$effort),
  family = delta_gamma(),
  time = "year", silent = FALSE,
  control = sdmTMBcontrol(newton_loops = 1)
)
tictoc::toc()
fit_sdmTMB
```

```{r compare-vast-sdmTMB1, echo = FALSE}
plot_betas <- function(vast_model, sdmTMB_model, vast_par = "beta1_ft", sdmTMB_pars = 1) {
  s <- vast_model$parameter_estimates$SD
  vast_est1 <- as.list(s, "Estimate", report = FALSE)
  vast_est2 <- as.list(s, "Estimate", report = TRUE)
  vast_sd1 <- as.list(s, "Std. Error", report = FALSE)
  vast_sd2 <- as.list(s, "Std. Error", report = TRUE)
  sdmTMB_est <- as.list(sdmTMB_model$sd_report, "Estimate", report = FALSE)
  sdmTMB_sd <- as.list(sdmTMB_model$sd_report, "Std. Error", report = FALSE)
  b_year_vast <- vast_est1[[vast_par]][!is.na(vast_sd1[[vast_par]])]
  b_year_vast_se <- vast_sd1[[vast_par]][!is.na(vast_sd1[[vast_par]])]
  years <- sort(unique(pcod$year))
  lwr_vast <- b_year_vast - 2 * b_year_vast_se
  upr_vast <- b_year_vast + 2 * b_year_vast_se
  plot(years, b_year_vast, ylim = range(c(lwr_vast, upr_vast)))
  segments(years, lwr_vast, years, upr_vast)
  years <- years + 0.05
  if (sdmTMB_pars == 1) {
    points(years, sdmTMB_est$b_j)
    segments(years, sdmTMB_est$b_j - 2 * sdmTMB_sd$b_j, 
    years, sdmTMB_est$b_j + 2 * sdmTMB_sd$b_j,
    col = "red")
  } else {
    points(years, sdmTMB_est$b_j2)
       segments(years, sdmTMB_est$b_j2 - 2 * sdmTMB_sd$b_j2, 
    years, sdmTMB_est$b_j2 + 2 * sdmTMB_sd$b_j2,
    col = "red")
  }
  legend("topright", legend = c("VAST", "sdmTMB"), 
    col = c("black", "red"), bty = "n", lty = c(1, 1))
}
```
